Lyman-$\alpha$ photons through rotating outflows

### 2018-04-01 paper/paper.tex
minor typos

However, the gas dynamics in Lyman Alpha Emitter (LAE) galaxies is
more complex than pure rotation.
The case for outflows as traced in the \lya line is much better
stablished.
In many observations the \lya line profiles has a single peak
redwards from the line's center, in other cases there is a double peak
but the peak on the red side is stronger
\citep[e.g.][]{2010ApJ...717..289S,Erb14,Trainor16}.   
These features been readily explained as the consequence of multiple
\lya photon scatterings through an homogeneous outflowing shell of
neutral Hydrogen
\citep{2006A&A...460..397V,Orsi12,2012ApJ...751...29Y,2015ApJ...812..123G}.

@@

\begin{figure*}
\centering
    \includegraphics[width=0.48\textwidth]{doppler_shift_logtau6_theta90}
    \includegraphics[width=0.48\textwidth]{doppler_shift_logtau6_theta0}
  \caption{\textbf{Qualitative trends of changing outflow and
      rotational velocity viewed perpendicular/paralell to the
      rotation axis}.  
    Here we fix $\tauh=10^6$. 
    The six panels on the left correspond to $\theta=90^\circ$ and the
    panels on the right to $\theta=0^{\circ}$
    We vary \vrot increasing from left to right and \vout increasing
    from top to bottom. 
    The thin black line corresponds to the \lya line obtained with
    CLARA without any rotation and the indicated outflow velocity.
    The thick black line corresponds to the results including both
    outflows and rotation.
    The thick gray line shows the results of modifyin the pure outflow
    solution by the Doppler shift presented in Equation \ref{eq:shift_x}
    (in thin line), if there is a radiative transfer of rotation and outflows
    (thick and clear line), and if there is a radiative transfer of
    only outflows, but also a Doppler shift from the rotational
    velocity (thick and dark line).  
    \label{fig:doppler_shift}}
\end{figure*}

@@

\begin{figure}
\centering
    \includegraphics[width=0.48\textwidth]{doppler}
  \caption{\textbf{Spectra from receeding/approaching sides of a toy
      model LAE.}  These results correspond to the RT simulation with
    $\vout=25$\kms, $\vrot=50$\kms, $\tauh=10^5$. 
    The spectra were computed for a viewing angle of
    $\theta=90^{\circ}$. This toy model illustrates to what extent spectra from
    opposite sides of a galaxy have an imprint of the rotational
    kinematics. 
    \label{fig:doppler}}
\end{figure}

@@

\begin{equation}
x' = x + \frac{\vec{v}_{\rm rot}\cdot\hat{k}}{v_{\rm th}}
\label{eq:shift_x}
\end{equation}
%
where $x'$ is the photon's new adimansional frequency, $x$ is the photon's
frequency after being procesed by the outflow, $v_{\rm rot}$ is the 
rotational velocity at the point of escape of the photon, $\hat{k}$ is
the photon's direction of propagation and $v_{\rm th}$ is the thermal
velocity of the sphere.

@@

The thick gray line correspods to the pure outflow solution
with the Doppler boost added to model rotation's influence.
At $\tauh=10^6$ the Doppler boost does a good job at capturing the broad
morphological features introduced by rotation: the angle dependence,
the broadening and the intensity increase at the line's center.

@@

\begin{equation}
\label{eq:bi}
\mathrm{BI} = \mathrm{KURTOSIS} - \mathrm{SKW}^2 = \frac{m_4}{m_2^{2}} - \frac{m_3^2}{m_2^{3}},
\end{equation}
%
where each $m_i$ is the i-th moment about the mean. 
The STD has velocity units and quantifies the line's width.
The SKW is adimensional and quantifies the peaks' asymmetry. 
In the case of a bimodal distribution, $SKW>0$ means that the blue
peak is taller and for $SWK<0$ the red peak is taller. 
The BI is adimensional and quantifies whether the line has 1 or 2
peaks: it is  always $\geq 1$ \citep{Pearson1929} and the 
to 1, the more bimodal is the line (i.e. has 2 similar peaks). 
We found by visual inspection of our spectra that BI$=2.5$ marks the
transition between two peaks (however imbalanced) and a dominang
single peak.

\subsubsection{Standard Deviation}
Figure \ref{fig:standard_deviation} summarizes the standard deviation
results for all our models.
Each panel shows the STD as a function of \vrot.
All panels were computed using a viewing angle of $\theta =
90^{\circ}$ (perpendicular to the rotation axis), which has the most
extreme ifluence from rotation.
 The black triangles
correspond to the full RT solution and the line to the DS
approximation.  
The optical depth increases from top to bottom and the outflow
velocity from left to right.
This quantitative plot confirms that the line width increases with
rotational velocity and optical depth.
These trends are expected; higher rotational velocities can be seen as
an addition of different Doppler shifts that smear out the line, while
a higher optical depth translates into a larger number of scatterings
that increase the probability of a photon to diffuse in frequency
resulting in a broader line.

The DS successfuly reproduces all trends with the optical depth,
rotational velocity and outflow velocity.
However, the DS consistently underestimates the STD. 
The difference between the RT and DS increases with the outflow
velocity and the rotational velocity, and decreases with increasing
optical depth.
In the range of parameter space explore, this difference has as an
upper bound of $\sim 7\%$, $3\%$ and $\sim 2\%$ for
 $\tauh=10^5$, $10^6$ and $10^7$, respectively.

@@

Figure \ref{fig:skewness} presents the skwewness results for all the
models together with the DS comparison following the same layout as
Figure \ref{fig:standard_deviation}.
In all cases the skewness is negative showing that all the lines
are unbalanced towards the red side of the spectrum.
Skewness increases with rotational velocity and decreases with
optical depth; rotation tries to smoothes the line diminishing the
asymetries while a higher optical depth reinforces the line asymetries.
The skewness does not have a monotonous trend with outflow velocity because
there is a transition between double and single peak line; for low
outflow velocities the skewness signals the balance between the two
existing peaks while for high outflow velocities it quantifies the
asymmetry of the already dominant read peak.

@@

The DS reproduces the main trends while underestimating the bimodality
statistics. 
As expected from the previous results the difference betwen RT and DS
decreases at higher optical depths and increases with increasing
values of the rotational and outflow velocities.
In this case the differences have an upper bound of $4\%$, $2\%$ and
$1\%$ for  $\tauh=10^5$, $10^6$ and $10^7$, respectively.

@@

In Figure \ref{fig:valley_intensity} we quantify how the intensity at
the line's center (i.e. the valley) changes with the viewing angle,
the outflow velocity and the optical depth.
These results correspond to a fixed rotational velocity of
\vrot$=100\ \kms$.
The triangles correspond to the RT simulations and the line represents
the DS results. 
The valley intensity is expresed as a fraction of the maximum peak
intensty in the line, as such the valley/peak ratio is always less
than one. 
In every panel we see that the valley/peak ratio decreases as the
observer moves from a line  of sight perpendicular to the rotation
axis onto a parallel line of sight. 
This is a clear demonstration of the viewing angle dependency
introduced by rotation.

@@

Another application of this new model to the current interpretation of \lya
spectra, is that the central ($V=0$) emission of the spectra seen in the
\lya line, is a consequence of the viewing angle of the galaxy and can be
controlled by it. 
Several authors have suggested that this central emission is caused by
radiation that escapes the galaxy without scattering, mostly due to a
clumpy interstellear medium \citep{Hansen06, 2016ApJ...833L..26G}.
Rotation is an alternative that solves this issue.

We note that the relative intensity of the two peaks in the outflowing
spectra is not modified by rotation, that asymmetry is solely
controled by the outflow velocity.
This means that results that have already derive a typical outflowing
velocity as to match the observational constraints do not have to be
completely revised. 
Only an additional exploration of rotational velocities and viewing
angles need to be explored.

A recent observational example is provided by
\cite{2017A&A...608L...4R}.
They report optical spectroscopy of a bright lensed galaxy at a
redshift of $z=2.4$. 
The object presents a triple peaked \lya profile that can be thought
as the superposition of two components, a double asymmetric peak and a
narrower symmetric peak around the line's center.
One possible option for the narrow component is that it escaped the
galaxy without substantial interaction with neutral Hydrogen.
We especulate that the central feature could also be due to a purely 
rotating compont that can produce such a feature as shown by
\cite{tololo}. The triple peak could be then produced by two main
decopled components: one rotating the other outflowing.

Another interesting example are the \lya lines in Green Pea galaxies \citep{2016ApJ...820..130Y}.
\cite{2017ApJ...844..171Y} showed in a study of 43 Green Pea galaxies
that 2/3 were strong \lya emitters. 
There the \lya escape fraction is effectively estimated from the ratio
of H$\alpha$ to \lya.
Galaxies with large escape fractions also substantial \lya emission at
the lines'center. 
We especulate that this feature could help to constrain rotation in
Green Pea galaxies.

@@

Although it is a good approximation to think the rotating spectra by a
sum of Doppler shifts, the peak of the spectra is also weighted by the
amount of mass with a given velocity. 
In this toy model the distance between the peaks of the
receeding/approaching spectra is close to $\sim 20\ \kms$, which is
less than half of the naively expected value of $2\vrot=100\ \kms$,
due to the fact that only a small fraction of the photons are emitted
at the extreme of the galaxy having the maximum rotational velocity of
$50\ \kms$.

@@

MWS provides public access to 831 spectra out of which 237 contain a
detected \lya line as dominant. 
The public dataset reports as the redshift of the galaxy the position
of the dominant peak.
We use this redshift to convert the spectra to a rest-frame and compute
the standard deviation and skewness statitstics. 
Given the uncertainty on the galaxy redshift we do not compute the
valley/peak ratio. 
We do not try to perform a full fit to all the spectra
\citep[e.g.][]{2017A&A...608A.139G} and wish only to
know whether the region of parameter space explored in our RT
simulations shows any crossing with the results from the MWS or
shows any sign of atypical features.

@@

To address the first question we posed in the introduction (\emph{ehat is then
the expected imprint of rotation on a resonant emission line such as
the \lya line?}) we can say that the main effects of rotation on the
\lya\ line morphology are:

@@

We find that these effects can be quantitatively explained by a
Doppler shift computed as product of quantities at the surface of last
scattering, namely  $\vec{v}_{\rm rot}\cdot \hat{k}$,  where
$\vec{v}_{\rm rot}$ is the velocity due to rotation and $\hat{k}$ is
the direction of the photon's propagation.  
These conclusions, specially the confirmation of the Doppler Boost as
a good approximation to model solid body rotation, strenghten the
semi-analytic model reported by \cite{Garavito14} when they modeled
the effect of pure rotation on \lya\ spectra.

@@

A more straightfoward approach takes into account how the
\lya spectra from two different sides of a galaxy can detect
approaching/receeding gas motions.
In that case one has to be careful and take into account that the
peaks of these two spectra is smaller than the expected value of
$2\times\vrot$ due to the differente weights at the surface of the
emitting regions.

@@

Our work confirms that rotation produces features in the \lya line
strong enough that can be expected to be detected in observations and
that should be taken into account into \lya modelling; the Doppler
Boost offers an easy-to-implement approximation to explore such
influence into already existing RT simulations and provide stronger
footing to the interpretation of \lya kinematics maps
\citep[e.g][]{2018MNRAS.473.3907A}.

>>>>

However, the gas dynamics in Lyman Alpha Emitter (LAE) galaxies is
more complex than pure rotation.
The case for outflows as traced in the \lya line is much better
established.
In many observations the \lya line profiles has a single peak
redwards from the line's center, in other cases there is a double peak
but the peak on the red side is stronger
\citep[e.g.][]{2010ApJ...717..289S,Erb14,Trainor16}.   
These features been readily explained as the consequence of multiple
\lya photon scatterings through an homogeneous outflowing shell of
neutral Hydrogen
\citep{2006A&A...460..397V,Orsi12,2012ApJ...751...29Y,2015ApJ...812..123G}.

@@

\begin{figure*}
\centering
    \includegraphics[width=0.48\textwidth]{doppler_shift_logtau6_theta90}
    \includegraphics[width=0.48\textwidth]{doppler_shift_logtau6_theta0}
  \caption{\textbf{Qualitative trends of changing outflow and
      rotational velocity viewed perpendicular/paralell to the
      rotation axis}.  
    Here we fix $\tauh=10^6$. 
    The six panels on the left correspond to $\theta=90^\circ$ and the
    panels on the right to $\theta=0^{\circ}$
    We vary \vrot increasing from left to right and \vout increasing
    from top to bottom. 
    The thin black line corresponds to the \lya line obtained with
    CLARA without any rotation and the indicated outflow velocity.
    The thick black line corresponds to the results including both
    outflows and rotation.
    The thick gray line shows the results of modifying the pure outflow
    solution by the Doppler shift presented in Equation \ref{eq:shift_x}
    (in thin line), if there is a radiative transfer of rotation and outflows
    (thick and clear line), and if there is a radiative transfer of
    only outflows, but also a Doppler shift from the rotational
    velocity (thick and dark line).  
    \label{fig:doppler_shift}}
\end{figure*}

@@

\begin{figure}
\centering
    \includegraphics[width=0.48\textwidth]{doppler}
  \caption{\textbf{Spectra from receding/approaching sides of a toy
      model LAE.}  These results correspond to the RT simulation with
    $\vout=25$\kms, $\vrot=50$\kms, $\tauh=10^5$. 
    The spectra were computed for a viewing angle of
    $\theta=90^{\circ}$. This toy model illustrates to what extent spectra from
    opposite sides of a galaxy have an imprint of the rotational
    kinematics. 
    \label{fig:doppler}}
\end{figure}

@@

\begin{equation}
x' = x + \frac{\vec{v}_{\rm rot}\cdot\hat{k}}{v_{\rm th}}
\label{eq:shift_x}
\end{equation}
%
where $x'$ is the photon's new adimensional frequency, $x$ is the photon's
frequency after being processed by the outflow, $v_{\rm rot}$ is the 
rotational velocity at the point of escape of the photon, $\hat{k}$ is
the photon's direction of propagation and $v_{\rm th}$ is the thermal
velocity of the sphere.

@@

The thick gray line corresponds to the pure outflow solution
with the Doppler boost added to model rotation's influence.
At $\tauh=10^6$ the Doppler boost does a good job at capturing the broad
morphological features introduced by rotation: the angle dependence,
the broadening and the intensity increase at the line's center.

@@

\begin{equation}
\label{eq:bi}
\mathrm{BI} = \mathrm{KURTOSIS} - \mathrm{SKW}^2 = \frac{m_4}{m_2^{2}} - \frac{m_3^2}{m_2^{3}},
\end{equation}
%
where each $m_i$ is the i-th moment about the mean. 
The STD has velocity units and quantifies the line's width.
The SKW is adimensional and quantifies the peaks' asymmetry. 
In the case of a bimodal distribution, $SKW>0$ means that the blue
peak is taller and for $SWK<0$ the red peak is taller. 
The BI is adimensional and quantifies whether the line has 1 or 2
peaks: it is  always $\geq 1$ \citep{Pearson1929} and the 
to 1, the more bimodal is the line (i.e. has 2 similar peaks). 
We found by visual inspection of our spectra that BI$=2.5$ marks the
transition between two peaks (however imbalanced) and a dominant
single peak.

\subsubsection{Standard Deviation}
Figure \ref{fig:standard_deviation} summarizes the standard deviation
results for all our models.
Each panel shows the STD as a function of \vrot.
All panels were computed using a viewing angle of $\theta =
90^{\circ}$ (perpendicular to the rotation axis), which has the most
extreme influence from rotation.
 The black triangles
correspond to the full RT solution and the line to the DS
approximation.  
The optical depth increases from top to bottom and the outflow
velocity from left to right.
This quantitative plot confirms that the line width increases with
rotational velocity and optical depth.
These trends are expected; higher rotational velocities can be seen as
an addition of different Doppler shifts that smear out the line, while
a higher optical depth translates into a larger number of scatterings
that increase the probability of a photon to diffuse in frequency
resulting in a broader line.

The DS successfully reproduces all trends with the optical depth,
rotational velocity and outflow velocity.
However, the DS consistently underestimates the STD. 
The difference between the RT and DS increases with the outflow
velocity and the rotational velocity, and decreases with increasing
optical depth.
In the range of parameter space explore, this difference has as an
upper bound of $\sim 7\%$, $3\%$ and $\sim 2\%$ for
 $\tauh=10^5$, $10^6$ and $10^7$, respectively.

@@

Figure \ref{fig:skewness} presents the skewness results for all the
models together with the DS comparison following the same layout as
Figure \ref{fig:standard_deviation}.
In all cases the skewness is negative showing that all the lines
are unbalanced towards the red side of the spectrum.
Skewness increases with rotational velocity and decreases with
optical depth; rotation tries to smooth the line diminishing the
asymmetries while a higher optical depth reinforces the line asymmetries.
The skewness does not have a monotonous trend with outflow velocity because
there is a transition between double and single peak line; for low
outflow velocities the skewness signals the balance between the two
existing peaks while for high outflow velocities it quantifies the
asymmetry of the already dominant read peak.

@@

The DS reproduces the main trends while underestimating the bimodality
statistics. 
As expected from the previous results the difference between RT and DS
decreases at higher optical depths and increases with increasing
values of the rotational and outflow velocities.
In this case the differences have an upper bound of $4\%$, $2\%$ and
$1\%$ for  $\tauh=10^5$, $10^6$ and $10^7$, respectively.

@@

In Figure \ref{fig:valley_intensity} we quantify how the intensity at
the line's center (i.e. the valley) changes with the viewing angle,
the outflow velocity and the optical depth.
These results correspond to a fixed rotational velocity of
\vrot$=100\ \kms$.
The triangles correspond to the RT simulations and the line represents
the DS results. 
The valley intensity is expressed as a fraction of the maximum peak
intensity in the line, as such the valley/peak ratio is always less
than one. 
In every panel we see that the valley/peak ratio decreases as the
observer moves from a line  of sight perpendicular to the rotation
axis onto a parallel line of sight. 
This is a clear demonstration of the viewing angle dependency
introduced by rotation.

@@

Another application of this new model to the current interpretation of \lya
spectra, is that the central ($V=0$) emission of the spectra seen in the
\lya line, is a consequence of the viewing angle of the galaxy and can be
controlled by it. 
Several authors have suggested that this central emission is caused by
radiation that escapes the galaxy without scattering, mostly due to a
clumpy interstellar medium \citep{Hansen06, 2016ApJ...833L..26G}.
Rotation is an alternative that solves this issue.

We note that the relative intensity of the two peaks in the outflowing
spectra is not modified by rotation, that asymmetry is solely
controlled by the outflow velocity.
This means that results that have already derive a typical outflowing
velocity as to match the observational constraints do not have to be
completely revised. 
Only an additional exploration of rotational velocities and viewing
angles need to be explored.

A recent observational example is provided by
\cite{2017A&A...608L...4R}.
They report optical spectroscopy of a bright lensed galaxy at a
redshift of $z=2.4$. 
The object presents a triple peaked \lya profile that can be thought
as the superposition of two components, a double asymmetric peak and a
narrower symmetric peak around the line's center.
One possible option for the narrow component is that it escaped the
galaxy without substantial interaction with neutral Hydrogen.
We speculate that the central feature could also be due to a purely 
rotating component that can produce such a feature as shown by
\cite{tololo}. The triple peak could be then produced by two main
decopled components: one rotating the other outflowing.

Another interesting example are the \lya lines in Green Pea galaxies \citep{2016ApJ...820..130Y}.
\cite{2017ApJ...844..171Y} showed in a study of 43 Green Pea galaxies
that 2/3 were strong \lya emitters. 
There the \lya escape fraction is effectively estimated from the ratio
of H$\alpha$ to \lya.
Galaxies with large escape fractions also substantial \lya emission at
the line's center. 
We speculate that this feature could help to constrain rotation in
Green Pea galaxies.

@@

Although it is a good approximation to think the rotating spectra by a
sum of Doppler shifts, the peak of the spectra is also weighted by the
amount of mass with a given velocity. 
In this toy model the distance between the peaks of the
receding/approaching spectra is close to $\sim 20\ \kms$, which is
less than half of the naively expected value of $2\vrot=100\ \kms$,
due to the fact that only a small fraction of the photons are emitted
at the extreme of the galaxy having the maximum rotational velocity of
$50\ \kms$.

@@

MWS provides public access to 831 spectra out of which 237 contain a
detected \lya line as dominant. 
The public dataset reports as the redshift of the galaxy the position
of the dominant peak.
We use this redshift to convert the spectra to a rest-frame and compute
the standard deviation and skewness statistics. 
Given the uncertainty on the galaxy redshift we do not compute the
valley/peak ratio. 
We do not try to perform a full fit to all the spectra
\citep[e.g.][]{2017A&A...608A.139G} and wish only to
know whether the region of parameter space explored in our RT
simulations shows any crossing with the results from the MWS or
shows any sign of atypical features.

@@

To address the first question we posed in the introduction (\emph{what is then
the expected imprint of rotation on a resonant emission line such as
the \lya line?}) we can say that the main effects of rotation on the
\lya\ line morphology are:

@@

We find that these effects can be quantitatively explained by a
Doppler shift computed as product of quantities at the surface of last
scattering, namely  $\vec{v}_{\rm rot}\cdot \hat{k}$,  where
$\vec{v}_{\rm rot}$ is the velocity due to rotation and $\hat{k}$ is
the direction of the photon's propagation.  
These conclusions, specially the confirmation of the Doppler Boost as
a good approximation to model solid body rotation, strengthen the
semi-analytic model reported by \cite{Garavito14} when they modeled
the effect of pure rotation on \lya\ spectra.

@@

A more straightforward approach takes into account how the
\lya spectra from two different sides of a galaxy can detect
approaching/receding gas motions.
In that case one has to be careful and take into account that the
peaks of these two spectra is smaller than the expected value of
$2\times\vrot$ due to the different weights at the surface of the
emitting regions.

@@

Our work confirms that rotation produces features in the \lya line
strong enough that can be expected to be detected in observations and
that should be taken into account into \lya modeling; the Doppler
Boost offers an easy-to-implement approximation to explore such
influence into already existing RT simulations and provide stronger
footing to the interpretation of \lya kinematics maps
\citep[e.g][]{2018MNRAS.473.3907A}.
